---
title: "Calculating uniqueness"
author: "Andrea Ballatore"
date: "27/03/2022"
output: html_document
---
# Setup 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(moments)
library(knitr)
library(corrr)
library(pastecs)
library(coop)
library(sf)
```

# Uniqueness functions

```{r uniq}

# FUNCTIONS ETC

range01 <- function(x){
  (x-min(x,na.rm = T))/(max(x,na.rm = T)-min(x,na.rm = T))
}


```

# Case studies

## Greater London (land use and Points of Interest)

```{r}

```

## Greater London (boroughs)

```{r}
lnd_boro = st_read('data/london/london_borough_profiles_2015/london_boroughs_profiles_2015.geojson')
# select variables
print(names(lnd_boro))
lnd_boro = lnd_boro %>% select(gss, name, population_density_per_hectare_2017, average_age_2017, pc_of_resident_population_born_abroad_2015, employment_rate_pc_2015, modelled_household_median_income_estimates_2012_13, proportion_of_seats_won_by_conservatives_in_2014_election, proportion_of_seats_won_by_labour_in_2014_election, median_house_price_2015) %>% rename(id = gss)
plot(lnd_boro %>% select(-id,-name), border='white')

# integrate missing data
# from https://data.gov.uk/dataset/248f5f04-23cf-4470-9216-0d0be9b877a8/london-borough-profiles-and-atlas/datafile/1e5d5226-a1a7-4097-afbf-d3bd39226c39/preview
#lnd_boro[lnd_boro$name == "Kensington and Chelsea", "gross_annual_pay_2016"] = 63620
```
### Prep data

Find problematic variables.

```{r}
# get correlations
lnd_boro_corr = lnd_boro %>% st_drop_geometry() %>% select(-name,-id) %>% correlate(method='kendall') %>% stretch()
# observe high correlations
lnd_boro_corr %>% filter(r > .5 | r < -.5)
# remove Labour variable
lnd_boro = lnd_boro %>% select(-proportion_of_seats_won_by_labour_in_2014_election)
```

```{r}
lnd_boro_desc_stats = t(stat.desc(lnd_boro %>% st_drop_geometry() %>% select(-id,-name), norm=T))
lnd_boro_desc_stats
```

```{r}
# histograms to see normality
ggplot(gather(lnd_boro %>% st_drop_geometry() %>% select(-name,-id)), aes(value)) + 
    geom_histogram(bins = 20) + 
    facet_wrap(~key, scales = 'free_x')
```
The only variable whose skewness is a concern is `median_house_price_2015`. It will be normalised with log10.

```{r}
# normalise and scale
lnd_boro$median_house_price_2015_log10 = log10(lnd_boro$median_house_price_2015+1)
lnd_boro = lnd_boro %>% select(-median_house_price_2015)

# skewness is now acceptable
skewness(lnd_boro$median_house_price_2015_log10)
```

Scale all numeric variables.

```{r}
lnd_boro_scaled = lnd_boro %>% mutate_if(is.numeric, scale) %>% mutate_if(is.numeric, range01)
lnd_boro_scaled %>% head(5)
```

Multidimensional scaling of boroughs.

```{r}
mds <- lnd_boro %>% select(-name,-id) %>% st_drop_geometry() %>%
  dist() %>% cmdscale() %>% as_tibble()
colnames(mds) <- c("Dim1", "Dim2")
mds$name = lnd_boro_scaled$name

# plot MDS
ggplot(mds, aes(x=Dim1, y=Dim2)) +
  geom_point() +
  geom_text(
    label=mds$name, 
    nudge_x = .25, nudge_y = .25, 
    check_overlap = T
  )
rm(mds)
```


### Uniqueness

Calculate uniqueness of boroughs based on all variables.

```{r}
n_var = ncol(lnd_boro_scaled %>% st_drop_geometry() %>% select(-name,-id))
sim_matrix = tcosine(lnd_boro_scaled %>% st_drop_geometry() %>% select(-name,-id)) %>% as_tibble()

# calculate cosine for cases with missing values
missing_cases = rowSums(is.na(sim_matrix))
missing_cases_row_idx = which(missing_cases == (ncol(sim_matrix)-1))
for (row_idx in missing_cases_row_idx){
  for (other_row_idx in seq(nrow(lnd_boro_scaled))){
    pair_cosine = tcosine(lnd_boro_scaled[c(row_idx,other_row_idx),] %>% st_drop_geometry() %>% select_if(is.numeric) %>% select_if(~ !any(is.na(.))))
    print(lnd_boro_scaled[c(row_idx, other_row_idx),])
    print(pair_cosine)
    TODO: save results
  }
}

lnd_boro_uniq = lnd_boro_scaled
lnd_boro_uniq$sim_sum = rowSums(sim_matrix, na.rm = T)-1
lnd_boro_uniq$sim_sum[lnd_boro_uniq$sim_sum == 0] <- NA
lnd_boro_uniq$uniq = lnd_boro_uniq$sim_sum / nrow(lnd_boro_uniq)

print(summary(lnd_boro_uniq$sim_sum))
print(summary(lnd_boro_uniq$uniq))
```


```{r}
lnd_boro_uniq %>% select(uniq) %>% plot(border = 'white')
```



## Countries (World Bank data)

```{r}

```

## EU (Eurostat data)


```{r}

```

End of notebook
